%
% file: dedication.tex
% author: V?ctor Bre?a-Medina
% description: Contains the text for thesis dedication
%

\chapter*{Dedication and acknowledgements}
\epigraph{\hspace{4ex}\textit{"Well, here at last, dear friends, on the shores of the Sea comes the end of our fellowship in Middle-earth. Go in peace! I will not say: do not weep, for not all tears are an evil."}}{--- J.R.R Tolkien,\\ \textit{The Return of the King}}

\begin{SingleSpace}
\initial{I} would like to thank Erik Meijer for all the great moments spent together hacking, laughing and discussing crazy theoretical topics on whiteboards and napkins. In the last few years you have taught me how to push my limits and not be afraid of anything, not only within our field, but in every challenge of my life. I'll always be grateful for that and I wish we'll get the chance to work together once again in the future.

I want to thank Georgios Gousios for all his help, support and honest feedback, as well as all my colleagues and friends at TU Delft for the great times and nerdy talks we shared. 

I want to thank Mircea, Georgi and Filippo, my brothers, for all these years spent together hacking, writing crazy Haskell derivations on a window, living the good life and most of all, doing it live!

I want to thank all my housemates and neighbors, the ones that are here and the ones that have left, you are the reason I can call this rainy and windy city of Delft my home. And of course, I want to thank you, my Pepek, for brightening up even the darkest of days with your tender smile.

Finally, my deepest thanks to my family for allowing me to chase my dreams so far from home, I love you all.

\begin{flushright}
\textit{Sincerely,}\\
\textit{Eddy Bertoluzzo}
\end{flushright}


\end{SingleSpace}
\clearpage